---
title: "MA369 Group Project #1"
author: "Andrea Marchi, Tristan Rumsey, Joey Scheuer, Rachel Hughes"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
setecho=T
knitr::opts_chunk$set(echo = setecho)
```


### Title Page and Individual Contributions
```{r}
#Part a - Lead: Andrea Marchi, Reviewer: Rachel Hughes
#Part b - Lead: Andrea Marchi, Reviewer: Rachel Hughes
#Part c - Lead: Andrea Marchi, Reviewer: Joey Scheuer
#Part d - Lead: Andrea Marchi, Reviewer: Joey Scheuer
#Part e - Lead: Andrea Marchi, Reviewer: Joey Scheuer
#Part f - Lead: Tristan Rumsey, Reviewer: Joey Scheuer
#Part g - Lead: Tristan Rumsey, Reviewer: Joey Scheuer
#Part h - Lead: Tristan Rumsey, Reviewer: Joey Scheuer
#Part i - Lead: Tristan Rumsey, Reviewer: Joey Scheuer
#Part j - Lead: Tristan Rumsey, Reviewer: Joey Scheuer
#Part k - Lead: Tristan Rumsey, Reviewer: Joey Scheuer




```


#Install needed dependencies
```{r, dependencies, echo=setecho}
# First thing you have to installing the following dependencies within library()
# and you only have to do this once
suppressWarnings({
  library(haven)
  library(readr)
  suppressMessages({library(GGally)})
})
```


## Part a
```{r, part_a, echo=setecho}
url = "https://stats.idre.ucla.edu/wp-content/uploads/2016/02/crime.sas7bdat"
crime_import = read_sas(url, NULL)
str(crime_import)
```
#### Answers and comments:
The dataset has a total of 51 observations, and a total of 9 variables. Of the 9 variables, 8 of them are numeric, and 1 is categorical (state).


## Part b
```{r, part_b, echo=setecho}
write.csv(crime_import, file = "crime_csv.csv", row.names = F)
write.table(crime_import, file = "crime_txt.txt", sep = "\t", row.names = F)
dir()
```
#### Answers and comments:
As seen above, the current working directory has a total of 7 different items, including the csv file of the data that we just created, titled "crime_csv", as well as the text file titled "crime_txt".


## Part c
```{r, part_c, echo=setecho}
crime_numeric = read_csv("crime_csv.csv",
                         col_types = cols(sid = col_skip(),
                                          state = col_skip(),
                                          crime = col_skip()))

str(crime_numeric)
```
#### Answers and comments:
We have removed the three variables called sid (state id), state, and crime, and assigned the remaining 6 variables to the "crime_numeric" dataframe. In the structure of this dataframe, there are 51 observations, and a total of 6 variables, all of which are numeric variables.


## Part d
```{r, part_d, echo=setecho}
boxplot(crime_numeric, las = 2, col = rainbow(ncol(crime_numeric)))
```
#### Answers and comments:
In the above boxplot, all 6 variables can be visualized in a different color. You can identify outliers for several of the variables. This includes 2 outliers for the "murder" variable, 2 outliers for the "pctwhite" variable, and 1 outlier for the "single" variable


## Part e
```{r, part_e, echo=setecho}
ggpairs(crime_numeric)
```
#### Answers and comments:
#Murder variable:
The murder variable has a heavily right-skewed distribution. 
#Pctmetro variable:
The pctmetro variable has a left-skewed distribution.
#Pctwhite variable:
The pctwhite variable has a heavily left-skewed distribution.
#Pcths variable:
The pcths variable has a relatively normal distribution.
#Poverty variable:
The poverty variable has a right-skewed distribution.
#Single variable:
The single variable has a heavily right-skewed distribution.
#Murder and pctmetro variables:
The murder variable has a weak positive association with pctmetro, as the correlation coefficient, r, equals 0.316, which is confirmed by the scatterplot that shows a positive slope.
#Murder and pctwhite variables:
The murder variable has a moderately strong negative association with pctwhite variable, as r = -0.706. This strong correlation is confirmed by the scatterplot that shows a strong relationship.
#Murder and pcths variables:
The murder variable has a moderately weak negative association with the pcths variable, as r = -0.286. This is confirmed by the scatterplot that shows a weak relationship.
#Murder and poverty variables:
The murder variable has a moderately strong positive association with the poverty variable, as r = 0.566. This is confirmed by the scatterplot that shows a moderately strong relationship.
#Murder and single variables:
The murder variable has a strong positive association with the single variable, as r = 0.859. This is confirmed by the scatterplot that shows a strong relationship.
#Pctmetro and pctwhite variables:
The pctmetro variable has a moderately weak negative association with the pctwhite variable, as r = -0.337. This is confirmed by the scatterplot that shows a moderately weak relationship.
#Pctmetro and pcths variables:
The pctmetro variable has a no association with the pcths variable, as r = -0.004. This is confirmed by the scatterplot that shows virtually no relationship.
#Pctmetro and poverty variables:
The pctmetro variable has a very weak negative association with the poverty variable, as r = -0.061. This is confirmed by the scatterplot that shows a very weak relationship.
#Pctmetro and single variables:
The pctmetro variable has a weak positive association with the poverty variable, as r = 0.260. This is confirmed by the scatterplot that shows a weak relationship.


## Part f
```{r, part_e, echo=setecho}

```
#### Answers and comments:



## Part g
```{r, part_e, echo=setecho}

```
#### Answers and comments:


## Part h
```{r, part_e, echo=setecho}

```
#### Answers and comments:


## Part i
```{r, part_e, echo=setecho}

```
#### Answers and comments:


## Part j
```{r, part_e, echo=setecho}

```
#### Answers and comments:


## Part k
```{r, part_e, echo=setecho}

```
#### Answers and comments:





